---
title: "Stat 579: Graphics with ggplot2"
author: "Heike Hofmann"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

# Outline

- Quick overview of a couple of useful R functions (building vocabulary)

- Intro to `ggplot2`

---

# Installing packages

Different ways to install packages, depending on where the package is published:

- package on CRAN (official archive):

```{r eval = FALSE}
install.packages("pkgname")
install.packages(c("package1", "package2", "package3"))
```

- package on github (less official, usually package under development)

```{r eval = FALSE}
remotes::install_github("github_handle/pkgname")
```

---
class: inverse
## Your Turn (5 min)

1. Install the package `classdata` from Dr Hofmann's github account (handle: `heike`)

2. Activate the package in your session and inspect the object `fbi`

3. What type of object is `fbi`? What are its dimensions? What is the time frame?

---

## Useful Object probing functions

for object `x`, the following commands:

- `x` return the object itself (not advisable for large objects)

- `head(x)` and `tail(x)` return the first\last six rows of data

- `summary(x)` provides a type-dependent summary of the object and its pieces (five-number summary for numeric variables, frequency break down for categorical)

- `str(x)` stands for *str*ucture, shows type of `x` and its parts

- `dim(x)` gives dimensionality of `x` (rows, columns)

- `names(x)` gives the name(s) of `x` and its sub-objects

---
class: inverse, middle, center
# Why `ggplot2`?

---

## Why `ggplot2`?

- Wildly popular package for statistical graphics: millions of downloads each month

```{r, echo=FALSE, cache = TRUE, fig.height = 5}
library(dlstats)
library(ggplot2)

dframe <- cran_stats("ggplot2")
ggplot(data = dframe[-nrow(dframe),], aes(x = end, y = downloads/1000)) + geom_point() + geom_line() +
  ylab("Number of monthly downloads (in thousands)") + xlab("Date") +
  ggtitle("Number of monthly downloads of ggplot2")
```

---

## `ggplot2`

- Developed by Hadley Wickham (An ISU Alumni)

- Designed to adhere to good graphical practices

- Constructs plots using the concept of layers

- Supports a wide variety plot types and extensions

<br>

- http://ggplot2.org/book/ or Hadley's book *ggplot2: Elegant Graphics for Data Analysis* for reference 

---

## Grammar of Graphics

A graphical representation (plot) consists of:

1. **mappings** (`aes`):  data variables are mapped to graphical elements

2. **layers**:  geometric elements (`geoms`, such as points, lines, 
rectangles, text, ...) and statistical transformations (`stats`, are identity, counts, bins, ...)

3. **scales**: map values in the data space to values in an aesthetic space (e.g. color, size, shape, but also position)

4. **coordinate system** (`coord`): normally Cartesian, but pie charts use e.g. polar coordinates

5. **facetting**: for small multiples (subsets) and their arrangement

6. **theme**: fine-tune display items, such as font and its size, color of background, margins, ...

---

## Scatterplots in `ggplot2`

`aes` allows us to specify mappings; scatterplots need a mapping for `x` and a mapping for `y`:

```
data(fbiwide, package = "classdata")

ggplot(data = fbiwide, aes(x = burglary, y = homicide)) +
  geom_point()
```  

```
ggplot(data = fbiwide, aes(x = log(burglary), y = log(homicide))) +
  geom_point()
```  

```
ggplot(data = fbiwide, aes(x = log(burglary), 
                           y = log(motor_vehicle_theft))) +
  geom_point()
```  

---

## The pipe operator `%>%`


`f(x) %>% g(y)` is equivalent to `g(f(x), y)`

i.e. the output of one function is used as input to the next function. This function can be the identity

Consequences:

- `x %>% f(y)` is the same as `f(x, y)`

- statements of the form `k(h(g(f(x, y), z), u), v, w)` become
`x %>% f(y) %>% g(z) %>% h(u) %>% k(v, w)`

- read `%>%` as "then do"

---

## Using the pipe `%>%`

```
ggplot(data = filter(fbi, Type=="homicide", 
aes(x = year, y = count)) + geom_point()
```

becomes

```
fbi %>% 
  filter(type=="homicide") %>%
  ggplot(aes(x = year, y = count)) + 
    geom_point()
```
---

## Aesthetics

Can map other variables to size or colour

```
ggplot(aes(x = log(burglary), y = log(motor_vehicle_theft),
           colour=state), data=fbiwide) + geom_point()

ggplot(aes(x = log(burglary), y = log(motor_vehicle_theft),
           colour=year), data=fbiwide) + geom_point()
```

```
ggplot(aes(x = log(burglary), y = log(motor_vehicle_theft),
           size=population), data=fbiwide) + geom_point()
```

other aesthetics: shape

---
class: inverse
## Your turn


- Work through each of the example plots

- Try variations of the plots to find answers to (some of) your questions. 

