---
title: "Stat 579: Visualizing Factor variables"
author: Heike Hofmann
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false

---

# Visualizing factors

- visualize factors directly: barcharts

- use factors in aesthetics (`colour`, `fill`, `shape`) or for facetting

- always make sure that the order in factors is sensible!


# Note: factors for fill color

In area plots (histograms and barcharts for now) use aesthetic `fill` for showing colored areas.

Only factor variables can be mapped to `fill`

---

# Example: 


```{r, message = FALSE, fig.height = 3}
library(classdata)
library(ggplot2)
ggplot(fbi, aes(x = Type, fill=Year)) + geom_bar() # no color
```

---

# Example (cont'd) 

Not sensible, but the colors show up:

```{r, fig.height = 3}
ggplot(fbi, aes(x = Type, fill=factor(Year))) + geom_bar() 
```


---

# Example: Survival on the titanic

The object `titanic` is a table of a break down of survival of passengers and crew on board the titanic by gender and age.

```{r, message = FALSE}
library(classdata)
head(titanic)
```


---

# Barcharts of all variables

.pull-left[
```{r, fig.width=3, fig.height = 2}
ggplot(titanic, aes(x = Survived)) +
  geom_bar()

ggplot(titanic, aes(x = Sex)) + 
  geom_bar()
```
]

.pull-right[
```{r, fig.width=3, fig.height = 2}
ggplot(titanic, aes(x = Class)) + 
  geom_bar()

ggplot(titanic, aes(x = Age)) + 
  geom_bar()
```
]

---

# `position="fill"`

.pull-left[
```{r, fig.width=3, fig.height = 2}
ggplot(titanic, aes(x = Survived, 
                    fill=Survived)) + 
  geom_bar(position="fill")

ggplot(titanic, aes(x = Sex, 
                    fill=Survived)) + 
  geom_bar(position="fill")
```
]

.pull-right[
```{r, fig.width=3, fig.height = 2}
ggplot(titanic, aes(x = Class, 
                    fill=Survived)) + 
  geom_bar(position="fill")

ggplot(titanic, aes(x = Age, 
                    fill=Survived)) + 
  geom_bar(position = "fill")
```
]

---

# Two and more factor variables

- besides facetting and position, use *mosaic plots*

- there are extension packages for `ggplot2`, e.g. `ggmosaic`

---

# Mosaicplots

```{r, fig.width=10, fig.height = 4, message=FALSE}
library(ggmosaic)
ggplot(data = titanic)  + 
  geom_mosaic(aes(x = product(Sex), fill=Survived, weight=1)) +
  facet_grid(Age~Class)
```

---
class: inverse
# Your turn 


For this your turn use the `titanic` object from the `classdata` package. (you might have to update the package using the command `remotes::install_github("heike/classdata")`) 

- Draw a barchart of Gender. Interpret.

- Map survival to fill color in the barchart of Gender. Interpret.

- In the previous barchart change the `position` parameter to `"fill"`. Interpret.

- Read up on the `position` parameter in `?geom_bar`. Try out other options for `position`.


---

# Why no pie charts?

## Order of tasks (easiest to hardest)

Cleveland and McGill (1984):

1. Position along a common scale
2. Positions along non-aligned scales
3. Length, direction, angle
6. Area
7. Volume, curvature
9. Shading, color saturation

Implications for drawing charts: make main comparisons as easy as possible (i.e. map as position along a common scale). Additional messages can be mapped on other aesthetics.
