---
title: "dplyr examples: mutate + group_by"
author: "Heike Hofmann"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: inverse, middle
# FBI data

---
class: inverse, middle
# `group_by` and `mutate`

---

# Working with `fbi`

`fbi` data included in the `classdata` package

upgrade the package: `remotes::install_github("heike/classdata")`

```{r, message=FALSE}
library(tidyverse)
library(classdata)
data("fbi", package="classdata")
tail(fbi)
```

---
class: inverse
# Your turn 

For this your turn use the `fbi` data from the `classdata` package

- Use `mutate` to introduce a variable `Rate` into the `fbi` data

- Use `mutate` to reorder types of crimes by (median) rate.

- Plot crime rates by type in side-by-side boxplots. Medians of the boxplots should be ordered.

---

#  `group_by` and `mutate` 

Introduce a ranking by rate for each type of crime.

```{r}
fbi <- fbi %>% group_by(Type) %>% mutate(
  Rate = Count/Population*70000,
  Rank = rank(Rate) # ranks from lowest rate to highest rate
)
fbi %>% filter(Rank == 1) %>% select(Type, State, Year, Rate)
```

---
class: inverse
# Your turn

For this your turn use the `fbi` data from the `classdata` package 
- Introduce a ranking by rate for states for each type of crime and year.

- Focus on the top states. Find a visual that shows how often each state managed to take the top spot since 1961.

```{r, echo=FALSE, fig.width=10, fig.height =5}
fbi <- fbi %>% group_by(Type, Year) %>% mutate(
  Rate = Count/Population*70000,
  Rank = rank(Rate) # ranks from lowest rate to highest rate
)
fbi %>% filter(Rank == 1) %>% 
  select(Type, State, Abb, Year, Rate) %>%
  ggplot(aes(x = Abb)) + 
  facet_wrap(~Type, scales="free", ncol=4) +
  geom_bar()

```
---

# Keywords: which function is it?


- `mutate`: introduce, replace, reorder, ...

- `summarise`: calculate, average, summary, ...

- `group_by`: for each, across, ... 

- `filter`: exclude, only consider, ...

---

# Potential traps

- using the $ notation in tidyverse can lead to strange behavior and error messages

- don't forget to save statements back into the dataset (`mutate`, `arrange`) or  new data objects (`summarise`, `filter`)

- when using the pipe `%>%`: what is output from lhs, first parameter on rhs?
