---
title: "Reshaping data: an example"
author: "Heike Hofmann"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: middle, inverse, center
# Attention: possibly sensitive

---
class: middle, inverse, center
# Daily covid testing at Iowa State

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=5, fig.height = 5)
library(tidyverse)
```

---

# The ISU website

ISU's public health website at https://asqk.ehs.iastate.edu/coviddashboard/TestingUpdateDaily/IsuTestingUpdateDashboardDaily
is updated three times a week. 

```{r, message=FALSE}
library(rvest)
url <- "https://asqk.ehs.iastate.edu/coviddashboard/TestingUpdateDaily/IsuTestingUpdateDashboardDaily"
root <- read_html(url)
tables <- root %>% html_table()
tables[[1]] %>% head()
```

---

# First overview

```{r}
testing <- tables[[1]]
testing %>% str()
```

---

# Basic cleaning

```{r}
testing <- testing %>% mutate(
  `Employee Positive` = 
    parse_number(`Employee Positive`, 
                 na=c("NA", "No testing conducted")),
  `Employee Negative` = 
    parse_number(`Employee Negative`, na=c("NA", "No testing conducted")),
  `Student Positive` = 
    parse_number(`Student Positive`, na=c("NA", "No testing conducted")),
  `Student Negative` = 
    parse_number(`Student Negative`, na=c("NA", "No testing conducted")),
  `Total Tested Positive` = 
    parse_number(`Total Tested Positive`, na=c("NA", "No testing conducted")),
  `Percent Positive` = 
    parse_number(`Percent Positive`, na=c("NA", "No testing conducted")),
  `Total Tested Negative` = 
    parse_number(`Total Tested Negative`, na=c("NA", "No testing conducted")),
  `Percent Negative` = 
    parse_number(`Percent Negative`, na=c("NA", "No testing conducted")),
  `Total Tested` = 
    parse_number(`Total Tested`, na=c("NA", "No testing conducted"))
)
```

---

# Second overview

```{r}
testing %>% str()
```

The first five variables are raw data, the others are derived from the first five

We keep the first five and re-derive the others when we have to:

```{r}
testing <- testing %>% select(1:5)
```

---

# Bring into long form

```{r}
library(tidyr)
tlong <- testing %>% 
  pivot_longer(cols = 2:5, 
               names_to = "Who Result", 
               values_to = "Number")
head(tlong)
```

---

# Separate into two columns

```{r}
tlong <- tlong %>% 
  separate(`Who Result`, 
           sep= " ", 
           into=c("Who", "Result"))
head(tlong)
```

---

# Make testing date a real date

```{r, message= FALSE}
library(lubridate)
tlong <- tlong %>% mutate(
  `Testing Date` = paste0(`Testing Date`, "/2020"),
  `Testing Date` = mdy(`Testing Date`)
)
head(tlong)
```
---

# Now we can analyze

```{r}
tlong %>% 
  ggplot(aes(x = `Testing Date`, weight=Number)) + 
  geom_bar() + 
  ylab("Number of total Tests")
```

---

```{r}
tlong %>% 
  filter(Result == "Positive") %>%
  ggplot(aes(x = `Testing Date`, weight=Number)) + 
  geom_bar() + 
  ylab("Number of positive Tests")
```

---

```{r, fig.width = 6}
tlong %>% 
  filter(Result == "Positive") %>%
  ggplot(aes(x = `Testing Date`, weight=Number)) + 
  geom_bar() + 
  ylab("Number of positive Tests") +
  facet_wrap(~Who)
```
