---
title: "Visualizing time"
author: "Heike Hofmann"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: middle, inverse, center
# Visualizing time

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
```

---

# NASA data

- `nasa` is a data set in the `GGally` package

- it consists of atmospheric measurements across a grid of locations in Middle America

```{r}
data(nasa, package="GGally")
head(nasa)
```

---

# Time series

- for each observational unit we have multiple measurements:

```{r, fig.width = 7, fig.height = 3}
library(tidyverse)
nasa %>% filter(x == 1, y == 1) %>%
  ggplot(aes(x = time, y = temperature)) + geom_point()
```

---

# Time series (2)

- for each observational unit we have multiple measurements,

- which we connect by a line:

```{r, fig.width = 7, fig.height = 3}
nasa %>% filter(x == 1, y == 1) %>%
  ggplot(aes(x = time, y = temperature)) + geom_line()
```

---

# Time series (3)

- for each observational unit we have multiple measurements,

- which we connect by a line.

- Each observational unit forms a group, we only connect points within a group by a line.

```{r, fig.width = 7, fig.height = 3}
nasa %>% filter(x == 1, y %in% c(1, 10)) %>%
  ggplot(aes(x = time, y = temperature, group=id)) + geom_line()
```

---
class: inverse
# Your Turn (6 mins)



- Load the `nasa` data from the package `GGally`,

- for one location, draw a time line of Ozone over the time frame (`time`).

- Plot separate lines for each of the years, i.e. put `month` on the x-axis and `ozone` on the y-axis for the same location. Is there a seasonal pattern apparent?


- Pick a location with x in 1:10 and y in 7:10. Plot temperature over time. Comment on the result.

