---
title: "Dates and times"
author: "Heike Hofmann"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: middle, inverse, center
# Working with date and time

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
```


---

# `lubridate` package

- package for working with dates and times

- defines different classes of time: instants, periods, intervals, durations

- defines converter and accessor functions, enables time calculus

---

# Converter functions and instants in time

- time instants: one (absolute) moment in time, e.g. `now()`, `Jan-1-2000`

- easy-to-use converter functions: 
    - date: `ymd`, `mdy`, `dmy`, ...
    - time: `hm`, `hms`, ...
    - date & time: `ymd_hms`, `mdy_hm`, ...

- order of letters determines how strings are parsed

- separators are automatically determined, then assumed to be the same

Example:
```{r message=FALSE}
mdy("11-10-2020")
```

---
class: inverse
# Your Turn 

- Create date objects for today's date by typing the date in text format and converting it with one of the `lubridate` converter functions. 

- Try different formats of writing the date and compare the end results.

---

# Example: ISU Police Crime Log

```{r, eval = FALSE}
# install java and run in terminal:
#  sudo R CMD javareconf
library(tabulizer)
library(dplyr)
library(ggplot2)
url <- "https://www.police.iastate.edu/crime-log/"
docs <- xml2::read_html(url)

location <- docs %>% rvest::html_nodes(".log") %>%
  rvest::html_attr(name="href")

tmp <- extract_tables(location)
final <- do.call(rbind, tmp[-length(tmp)])

headers <- c('Case.Number', 'Classification', 'Date.Reported', 'Time.Reported', 
             'Earliest.Occurrence.Date', 'Earliest.Occurrence.Time',
             'Latest.Occurrence.Date', 'Latest.Occurrence.Time',
             'General Location', 'Disposition')

final <- data.frame(final)
names(final) <- headers
head(final)
final <- final[-1,]
write.csv(final, "Crime-log-Nov05.csv", row.names=FALSE)
```


---

# Example: ISU Police Crime Log

```{r, message=FALSE}
url <- "https://raw.githubusercontent.com/Stat579-at-ISU/materials/master/04_charts/Crime-log-Nov05.csv"
isu <- read.csv(url, stringsAsFactors = FALSE)
str(isu)
```

---

# Example: ISU Police Crime Log

Convert date variables to actual dates:

```{r}
isu <- isu %>% mutate(
  Date.Reported = mdy(Date.Reported),
  Earliest.Occurrence.Date = mdy(Earliest.Occurrence.Date),
  Latest.Occurrence.Date = mdy(Latest.Occurrence.Date)
)

summary(isu)
```
---

# Example: ISU Police Crime Log

Oct 11 has 8 incidences?

```{r}
isu %>% 
  group_by(Date.Reported) %>%
  summarize(n = n()) %>%
  ggplot(aes(x = Date.Reported, y = n)) + geom_point()
```

---

# Example: ISU Police Crime Log 

What happened on Oct 11?

```{r}
isu %>% filter(Date.Reported== ymd("2020/10/11"))
```

What happened on Oct 10?

ISU Cyclones beat Texas Tech 31 to 15 (at Jack Trice)

---

# Accessor functions


- accessor functions: `year`, `month`, `week`, `wday`, `mday`, `yday`, `hour`, `minute`, ...

- accessor functions can also be used for setting elements of date and time, e.g. `hour(x) <- 12`

Examples:

```{r}
month(now())
wday(now(), label = TRUE)
```

---
class: inverse
# Your Turn: ISU Police (cont'd)

Use the `isu` data (download from class website or download directly) to answer the following questions:

- How many incidences did ISU police report on Oct 10? Where did those occur?
- Can you identify a  pattern for the week? Draw a barchart.
- How many different Mondays, Tuesdays, Wednesdays ... are in the data?
- On how many days were there no reports?

Challenge: What is the average number of incidences? Hint look into `?tiydr::complete`

---
class: inverse
# Your Turn: ISU Police (cont'd)


Use the `isu` data (download from class website or download directly) to execute the following tasks:

- times in the `isu` data set are reported in the form of 1 to 4 digit integers describing hour and minute of the day. For the variable `Time.Reported` separate these values into hour and minute variables.
Plot a barchart of the hour in which incidences are reported.

- accessor functions such as `hour(.)` and `minute(.)` can be used on the left hand side of an assignment.
Introduce a variable Date.Time.Reported into the dataset, that captures both time and date of an incidence report.


---

# Intervals and Durations

- Intervals have a *start* and an *end* date/time: absolute difference

- Durations are potentially of relative length (months, leap year, leap second, ...)

```{r}
end_date <- now()
# span is interval, years(1) is duration
span <- end_date - years(1)
span
end_date - days(10)
```


