---
title: "Dates and times"
author: "Heike Hofmann"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: middle, inverse, center
# Working with date and time

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
```


---

# `lubridate` package

- package for working with dates and times

- defines different classes of time: instants, periods, intervals, durations

- defines converter and accessor functions, enables time calculus

---

# Converter functions and instants in time

- time instants: one (absolute) moment in time, e.g. `now()`, `Jan-1-2000`

- easy-to-use converter functions: 
    - date: `ymd`, `mdy`, `dmy`, ...
    - time: `hm`, `hms`, ...
    - date & time: `ymd_hms`, `mdy_hm`, ...

- order of letters determines how strings are parsed

- separators are automatically determined, then assumed to be the same

Example:
```{r message=FALSE}
mdy("11-15-2018")
```

---
class: inverse
# Your Turn 

- Create date objects for today's date by typing the date in text format and converting it with one of the `lubridate` converter functions. 

- Try different formats of writing the date and compare the end results.


---

# Accessor functions


- accessor functions: `year`, `month`, `week`, `wday`, `mday`, `yday`, `hour`, `minute`, ...

- accessor functions can also be used for setting elements of date and time, e.g. `hour(x) <- 12`

Examples:

```{r}
month(now())
wday(now(), label = TRUE)
```

---

# Intervals and Durations

- Intervals have a *start* and an *end* date/time: absolute difference

- Durations are potentially of relative length (months, leap year, leap second, ...)

```{r}
end_date <- now()
# span is interval, years(1) is duration
span <- end_date - years(1)
span
end_date - days(10)
```

---

# Example: Movies

```{r, message=FALSE}
library(classdata)
summary(box$Date) # date variable: allows date calculus
```


---

# Example: Movies (cont'd)

```{r, fig.height = 3.25}
library(tidyverse)

box %>% ggplot(aes(x = Date)) + geom_histogram(binwidth=7)
```

---

# Example: Movies (cont'd)

Is there a seasonal effect in the number of movies in the box office?

```{r, fig.height = 3.25}
library(lubridate)

box %>% ggplot(aes(x = month(Date, label=TRUE))) + geom_bar()
```

---
class: inverse
# Your Turn (10 min)

- inspect the `budget` data set from the `classdata` package,

- make sure the variable `Release Date` is a date format,

- plot a histogram of the variable. Set the bin width appropriately.

- is there a monthly trend apparent for movies that were released since 2000? draw barcharts of month, facet by year. What about day of the week?



