---
title: "Functions in R"
author: "Heike Hofmann"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: middle, inverse, center
# Function writing

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
```

---

# Functions in R

- Have been using functions a lot, now we want to write them ourselves!

- Idea: avoid repetitive coding (errors will creep in)

- Instead: extract common core, wrap it in a function, make it reusable


---

# Structure of functions

- Name

- Input arguments

    - names, 

    - default values

- Body

- Output values

---

# A first function

```{r}
mymean <- function(x) {
	return(sum(x)/length(x))
}
```

```{r}
mymean(1:15)
mymean(c(1:15, NA))
```

---

# A first function (2)

```{r}
mymean <- function(x, na.rm=F) {
	if (na.rm) x <- na.omit(x)
	
	return(sum(x)/length(x))
}

mymean(1:15)
mymean(c(1:15, NA), na.rm=T)
```


---
class: inverse
# Your Turn 

The-Numbers website gives an overview of weekly boxoffice gross at https://www.the-numbers.com/weekend-box-office-chart

The package `rvest` allows us to download data from a website with the url `url` using the following lines of code.

```
library(rvest)
doc <- read_html(url)
tables <- doc %>% html_table()

head(tables[[2]])
```

Write a function that uses the url as input argument, scrapes the data, cleans it up, and returns the cleaned data.

---

# Your turn - solution

```{r warning = FALSE, message = FALSE}
url <- "https://www.the-numbers.com/weekend-box-office-chart"

boxoffice_scraper <- function(url) {
  require(rvest)
  require(readr)
  html <- read_html(url)
  tables <- html %>% html_table(fill=TRUE)
  box <- tables[[2]]
  names(box)[1:2] <- c("Rank", "Rank.Last.Week")
  box <- box %>% mutate(
    Gross = parse_number(Gross),
    Thr = parse_number(Thr)
  )
  box  
}
```

---

# Now try it out

```{r, eval=TRUE, cache =TRUE, message=FALSE}
box <- boxoffice_scraper("https://www.the-numbers.com/weekend-box-office-chart")
head(box)
```

---
class: inverse, center, middle
# Always scrape data responsibly!
